Mega-analysis of the Interoceptive Accuracy Scale (IAS) Structure and its Subjective Correlates

### 2025-10-17 paper/1_Initial/manuscript.qmd
changes to manuscript

# General Discussion

@@

A short version of the IAS has been proposed before. @lin2023 suggested a 12-item version by removing items with local dependency, high difficulty, and/or low discrimination. Their parallel analysis suggested unidimensionality, although this method is known to underestimate factor numbers in multidimensional data [@golino2017exploratory; @markos2024dimensionality]. In contrast, our study applied multiple methods such as EGA, FA, and HCA yet the same pairs consistently emerged across datasets and demonstrated superior fit compared to alternative models. This challenges the assumption of unidimensionality of the IAS [@murphy2019; @brand2023]. Although the short version of the IAS by @lin2023 preferred a one-factor solution, their 12-item IAS was derived by removing locally dependent items identified through Item Response Theory (IRT). In our framework, these locally dependent items are instead interpreted as meaningful pairs (e.g., Heart–Breath), reflecting distinct bodily domains. However, because IRT assumes unidimensionality [@nguyen2014introduction], it may have treated these natural item clusters as sources of statistical dependency rather than as theoretically relevant structures within interoceptive accuracy.

<!-- Study 2 findings -->
Study 2 provided further evidence for a domain-specific perspective. Certain item pairings, particularly Hunger–Thirst and Bruise–Blood sugar, showed consistent negative correlations with depression and anxiety measures, in some cases approaching or even surpassing the strength of associations observed for the full IAS scale. Notably, these were also the only pairs, out of the seven examined, that displayed significant positive correlations with the BPQ Body Awareness, whereas the full IAS scale showed no such association. This pattern suggests that specific item pairings may show distinct patterns of association with other interoceptive measures and with mental health indices, whereas the aggregated IAS scale may obscure these relationships. This highlights that questionnaires like the IAS may capture multiple facets of interoception rather than a single underlying dimension, and that domain-specific approaches may provide greater construct validity.

These findings mirror broader inconsistencies in the literature concerning associations between the IAS and the BPQ Body Awareness. While some studies report significant negative associations [@lin2023], others have observed significant positive associations [@brand2023; @campos2021; @gaggero2021; @todd2022], and still others none at all [@murphy2019]. In our study the full IAS was not associated with the BPQ Body Awareness. Interpretation is further complicated by conceptual debates: the BPQ Body Awareness is frequently treated as a measure of interoception, whereas the BPQ Autonomic Reactivity subscale has been criticized as indexing dysfunction in the nervous system rather than interoceptive sensitivity [@gaggero2021]. Our results lend tentative support to this interpretation, as the full IAS scale and most item pairings showed significant negative associations with the BPQ Autonomic Reactivity, with the exception of the Bruise–Blood sugar pair. This pattern is consistent with the idea that this subscale may capture maladaptive bodily experiences or autonomic dysfunction, rather than interoceptive accuracy per se. However, the Body Awareness subscale of the BPQ has also been criticized for its ambiguous instructions. It is often unclear whether respondents should interpret the items as assessing attentional focus on bodily sensations, perceived accuracy in detecting them, or simply the frequency of experiencing such signals. As a result, studies have reported that observed associations vary depending on how the BPQ Body Awareness is constructed [@campos2021; @gabriele2022dissociations; @gaggero2021]. As such, it is unclear whether the lack of a relationship between the IAS and the BPQ Body Awareness in this study reflects a genuine divergence between the constructs measured by the two scales, or instead arises from ambiguities in how the BPQ Body Awareness is operationalized and interpreted across studies.

Taken together with these debates, the discrepancies observed here highlight that some IAS pairings may capture interoceptive self-reported accuracy more effectively than the aggregated measure, raising questions about the interpretability of the full scale. By contrast, other pairings (e.g., Cough–Sneeze, Wind–Burp) were weakly or inconsistently related to external variables, raising concerns about their usefulness. Together, these findings indicate that the strength and relevance of interoceptive self-reported accuracy may vary systematically across physiological domains. At the same time, the interpretation of such associations is complicated by the lack of contextualisation in the items themselves.For example, a racing heart is often associated with anxiety, yet the same sensation can also arise during exercise or other forms of physiological arousal. Without contextual anchors, it remains unclear which bodily states participants are drawing upon when responding. In line with prior work emphasizing the importance of contextual factors in the measurement of interoception [@vlemincx2023novel], our results suggest that the value of self-report lies less in broad aggregation across domains and more in the targeted assessment of physiologically and situationally meaningful contexts.

<!-- Correlates of IAS Across Psychological Domains -->
The dispositional correlates of the IAS help delineate the conceptual boundaries of interoceptive accuracy as a construct. Self-reported interoceptive accuracy, as assessed by the IAS, has been robustly associated with mood-related outcomes, particularly depression and anxiety, underscoring its relevance for affective regulation and bodily state monitoring (Benau et al., 2023; Brand et al., 2023). This aligns with a broader body of research linking affective conditions such as anxiety and depression with altered interoceptive processing (e.g., Garfinkel et al., 2016). However, recent findings have also reported null or inconsistent associations between interoceptive abilities and mental health symptomatology, suggesting that such relationships may not generalize across all interoceptive domains or measurement modalities [@banellis2025interoceptive; @jenkinson2024interoception]. For instance, a recent review by @jenkinson2024interoception found no consistent evidence linking cardiac interoceptive accuracy tasks with anxiety or depression across multiple meta-analyses. Collectively, these findings underscore the importance of examining interoceptive self-perception at the level of distinct bodily domains rather than relying on aggregated or unidimensional measures.

In contrast, associations of IAS scores with personality traits and higher-order beliefs were modest: small positive correlations emerged with Openness and Agreeableness, and weak associations with primal world beliefs. Consistent with recent work linking interoceptive self-report to personality dispositions [@haustein2024interoception; @simkute2025interoception], these results suggest that interoceptive accuracy primarily operates at the level of bodily self-perception. This contrast reinforces the view that the IAS primarily indexes bodily self-perception, displaying only modest associations with  broader dispositional traits.

A potential key reason for these discrepancies is the lack of a consistent theoretical framework for interoception, as dimensions are neither defined nor captured equally across questionnaires and experimental tasks [@desmedt2022measures]. **[complete paragraph]**

<!-- IAS limitations and potential improvements -->
These two studies revealed three main points. Firstly, a shortened, 14-item refined IAS retained greater psychometric integrity while removing unstable and/or redundant items. Secondly, item pairings combined with analogue-style ratings captured greater variability and allowed more precise differentiation among individuals, often outperforming global scores in predictive value. However, even the refined IAS has significant limitations. Although the 14-item version improves interpretability and efficiency, the scale continues to underrepresent key interoceptive modalities such as cardiac, respiratory, and nuanced somatosensory signals, limiting its ability to capture the full breadth of interoceptive processes. This shortcoming underscores the need for broader coverage in future scale development. Nevertheless, analogue-style items and item pairings offer clear advantages: they capture greater response variability, allow more precise differentiation among individuals, and provide efficient, domain-specific indicators that often outperform global scores in predictive value. Building on these insights, a next-generation interoception scale could balance breadth (multiple items per modality) with depth (domain-specific subscales), retain the benefits of analogue or wider rating ranges (e.g., 1–7 instead of 1–5), and optimize coverage across physiologically and psychologically relevant bodily domains. Such a scale would improve the interpretability, sensitivity, and practical utility of interoceptive self-report measures, ultimately allowing for a more nuanced mapping of interoceptive abilities to affective, cognitive, and health-related outcomes.

@@

Because the IAS is a newly developed questionnaire and the only one designed to assess interoceptive self-reported accuracy - a dimension previously thought to require objective tasks such as the HDT and HCT - our focus was on evaluating its structure and correlates without introducing established measures with documented shortcomings. Nonetheless, future research should employ novel methods, such as Exploratory Graph Analysis, to clarify the convergent and divergent validity of these questionnaires across interoceptive domains.

<!-- Future Outlook in the field -->
Building on these limitations, a next-generation interoception scale should balance breadth (multiple items per modality, including underrepresented domains such as cardiac and respiratory) with depth (domain-specific subscales). Retaining analogue or wider rating ranges (e.g., 1–7 instead of 1–5) would further enhance sensitivity and coverage. Such a scale would improve the interpretability, sensitivity, and practical utility of interoceptive self-report measures, ultimately allowing for a more nuanced mapping of interoceptive abilities to affective, cognitive, and health-related outcomes.

<!-- Conclusion -->
Taken together, these studies advance understanding of the IAS and interoceptive self-report more broadly. The refined 14-item IAS provides a more efficient and interpretable measure, capturing domain-specific interoceptive signals that are differentially associated with psychological outcomes. Consistent associations with mood, alongside modest links to personality and beliefs, indicate that the IAS reflects adaptive, context-sensitive aspects of bodily self-perception rather than broad dispositional traits. At the same time, current findings highlight limitations in modality coverage and item specificity, suggesting that future scale development should incorporate multimodal, context-sensitive, and continuous analogue ratings or wider discrete scales to better capture the richness of interoceptive processes and clarify their role in affective, cognitive, and health-related outcomes [@desmedt2022measures; @vlemincx2023novel]. Ultimately, these studies show that the IAS is most informative when treated as a set of domain-specific, context-sensitive indicators rather than as a broad global index, advancing both the measurement and theoretical understanding of interoception.

>>>>

# General DiscussionGeneral Discussion

@@

A short version of the IAS has been proposed before.  @lin2023 suggested a 12-item version by removing items with local dependency, high difficulty, and/or low discrimination. Their parallel analysis suggested unidimensionality, although this method is known to underestimate factor numbers in multidimensional data [@golino2017exploratory; @markos2024dimensionality]. In contrast, our study applied multiple methods such as EGA, FA, and HCA yet the same pairs consistently emerged across datasets and demonstrated superior fit compared to alternative models. This challenges the assumption of unidimensionality of the IAS [@murphy2019; @brand2023]. Although the short version of the IAS by  @lin2023 preferred a one-factor solution, their 12-item IAS was derived by removing locally dependent items identified through Item Response Theory. In our framework, these locally dependent items are instead interpreted as meaningful pairs (e.g., Heart–Breath), reflecting distinct bodily domains. However, because IRT assumes unidimensionality [@nguyen2014introduction], it may have treated these natural item clusters as sources of statistical dependency rather than as theoretically relevant structures within interoceptive accuracy.

<!-- Study 2 findings -->
Study 2 provided further evidence for a domain-specific perspective. Certain item pairings, particularly Hunger–Thirst, Bruise–Blood sugar and Muscles-Pain, showed consistent negative correlations with depression and anxiety measures, in some cases approaching or even surpassing the strength of associations observed for the full IAS scale. Notably, Hunger-Thirst and Bruise-Blood sugar were also the only pairs, out of the seven examined, that displayed significant positive correlations with the BPQ Body Awareness, whereas the full IAS scale showed no such association. This pattern suggests that specific item pairings may show distinct patterns of association with other interoceptive measures and with mental health indices, whereas the aggregated IAS scale may obscure these relationships. This highlights that questionnaires like the IAS may capture multiple facets of interoception rather than a single underlying dimension, and that domain-specific approaches may provide greater construct validity.

These findings mirror broader inconsistencies in the literature concerning associations between the IAS and the BPQ Body Awareness. While some studies report significant negative associations [@lin2023], others have observed significant positive associations  [@brand2023; @campos2021; @gaggero2021; @todd2022], and still others none at all  [@murphy2019]. In our study the full IAS was not associated with the BPQ Body Awareness. Interpretation is further complicated by conceptual debates: the BPQ Body Awareness is frequently treated as a measure of interoception, whereas the BPQ Autonomic Reactivity subscale has been criticized as indexing dysfunction in the nervous system rather than interoceptive sensitivity [@gaggero2021]. Our results lend tentative support to this interpretation, as the full IAS scale and most item pairings showed significant negative associations with the BPQ Autonomic Reactivity, with the exception of the Bruise–Blood sugar pair. This pattern is consistent with the idea that this subscale may capture maladaptive bodily experiences or autonomic dysfunction, rather than interoceptive accuracy per se. However, the Body Awareness subscale of the BPQ has also been criticized for its ambiguous instructions. It is often unclear whether respondents should interpret the items as assessing attentional focus on bodily sensations, perceived accuracy in detecting them, or simply the frequency of experiencing such signals. As a result, studies have reported that observed associations vary depending on how the BPQ Body Awareness is interpreted [@campos2021; @gabriele2022dissociations; @gaggero2021]. As such, it is unclear whether the lack of a relationship between the IAS and the BPQ Body Awareness in this study reflects a genuine divergence between the constructs measured by the two scales, or instead arises from ambiguities in how the BPQ Body Awareness is operationalized and interpreted across studies.

Taken together with these debates, the discrepancies observed here highlight that some IAS pairings may capture interoceptive self-reported accuracy more effectively than the aggregated measure, raising questions about the interpretability of the full scale. By contrast, other pairings (e.g., Cough–Sneeze, Wind–Burp) were weakly or inconsistently related to external variables, raising concerns about their usefulness. Together, these findings indicate that the strength and relevance of interoceptive self-reported accuracy may vary systematically across physiological domains. At the same time, the interpretation of such associations is complicated by the lack of contextualisation in the items themselves.For example, a racing heart is often associated with anxiety, yet the same sensation can also arise during exercise or other forms of physiological arousal. Without contextual anchors, it remains unclear which bodily states participants are drawing upon when responding, highlighting the interpretive ambiguity of self-reports. In line with prior work emphasizing the importance of contextual factors in the measurement of interoception [@vlemincx2023novel], our results suggest that the value of self-report lies less in broad aggregation across domains and more in the targeted assessment of physiologically and situationally meaningful contexts.

<!-- Correlates of IAS Across Psychological Domains -->
The dispositional correlates of the IAS help delineate the conceptual boundaries of interoceptive accuracy as a construct. Self-reported interoceptive accuracy, as assessed by the IAS, has been robustly associated with mood-related outcomes, particularly depression and anxiety, underscoring its relevance for affective regulation and bodily state monitoring [@benau2023self; @brand2023]. This aligns with a broader body of research linking affective conditions such as anxiety and depression with altered interoceptive processing [e.g.,@garfinkel2015knowing]. However, recent findings have also reported null or inconsistent associations between interoceptive abilities and mental health symptomatology, suggesting that such relationships may not generalize across all interoceptive domains or measurement modalities (@banellis2025interoceptive; @jenkinson2024interoception). For instance, a recent review by @jenkinson2024interoception found no consistent evidence linking cardiac interoceptive accuracy tasks with anxiety or depression across multiple meta-analyses.Collectively, these findings indicate that relationships between interoceptive accuracy and mental health may vary across bodily domains and measurement modalities. This suggests that the IAS, while informative, may benefit from a more differentiated approach that accounts for domain-specific and contextual influences on self-reported interoceptive accuracy.

In contrast to the relatively stronger associations observed with affective measures, correlations between IAS scores and personality traits were small but directionally consistent. Modest positive associations emerged with Openness, Agreeableness, and Conscientiousness, while Neuroticism and Extraversion showed near-zero relationships, and Honesty–Humility displayed weak negative correlations. This pattern suggests that self-reported interoceptive accuracy, as indexed by the IAS, shares limited but systematic overlap with broad dispositional tendencies -particularly those reflecting intellectual curiosity and creativity, patience and tolerance in interpersonal interactions, and diligence and organizational tendencies [@aashton2007empirical]. Recent evidence supports the view that relationships between interoceptive measures and personality are multifaceted rather than uniform. For instance, @haustein2024interoception reported significant associations between interoceptive awareness (MAIA) and all personality traits except Agreeableness, with Openness showing positive correlations with Body Listening and Emotional Awareness, and Neuroticism showing negative correlations with Self- and Attention Regulation. Similarly, @simkute2025interoception found that task-based indices of interoceptive sensitivity related negatively to Extraversion. Collectively, these findings indicate that interoceptive constructs show distinct and sometimes opposing patterns of association across measurement modalities, suggesting that the IAS relates to a specific, self-evaluative aspect of interoceptive functioning rather than reflecting a general interoceptive or personality-based trait.

A potential key reason for discrepancies in observed associations, including those seen with the IAS, is that self-report questionnaires in interoception often tap into distinct constructs rather than a common underlying dimension [@desmedt2022measures]. Measures labeled as interoceptive sensibility, awareness, or self-reported accuracy capture partially overlapping but empirically distinct facets, with low intercorrelations between questionnaires and items. Consequently, differences in associations with psychological outcomes or personality traits may reflect differences in what each questionnaire assesses rather than true variation in interoceptive functioning.

<!-- IAS limitations and potential improvements -->
These two studies revealed three main points. First,  in the datasets that included analogue-style ratings, these scales captured greater variability, further enhancing the ability to detect individual differences. Second, a shortened, 14-item version of the IAS retained strong psychometric integrity while removing unstable or redundant items. Third, specific item pairings allowed more precise differentiation among individuals and often outperformed global scores in predictive value. Nevertheless, even this refined IAS has important limitations. While the 14-item version improves interpretability and efficiency, it continues to underrepresent key interoceptive modalities such as cardiac, respiratory, and nuanced somatosensory signals, limiting its ability to capture the full breadth of interoceptive processes. This shortcoming underscores the need for broader coverage in future scale development. At the same time, analogue-style items and item pairings offer clear advantages: they increase response variability, allow more precise differentiation among individuals, and provide domain-specific indicators that often outperform global scores. Building on these insights, a next-generation interoception scale could balance breadth (multiple items per modality) with depth (domain-specific subscales), retain the benefits of analogue or wider rating ranges (e.g., 1–7 instead of 1–5), and optimize coverage across physiologically and psychologically relevant bodily domains. Such a scale would improve interpretability, sensitivity, and practical utility, ultimately enabling a more nuanced mapping of interoceptive abilities onto affective, cognitive, and health-related outcomes.

@@

Because the IAS is a newly developed questionnaire and the only one designed to assess interoceptive self-reported accuracy - a dimension previously assessed solely through objective tasks such as the HDT and HCT - our focus was on evaluating its structure and correlates without introducing established measures with documented shortcomings. Nonetheless, future research should employ novel methods, such as Exploratory Graph Analysis, to clarify the convergent and divergent validity of these questionnaires across interoceptive domains.

<!-- Conclusion -->
Taken together, these studies advance understanding of the IAS and interoceptive self-report more broadly. The refined 14-item IAS provides a more efficient and interpretable measure, capturing domain-specific interoceptive signals that show differential associations with psychological outcomes. Associations were consistently stronger with mood measures and compared to personality and beliefs, indicating that self-reported interoceptive accuracy relates more closely to affective experience than to stable dispositional traits. At the same time, current findings highlight limitations in modality coverage and item specificity, suggesting that future scale development should incorporate multimodal, context-sensitive, and continuous analogue ratings or wider discrete scales to better capture the richness of interoceptive processes and clarify their role in affective, cognitive, and health-related outcomes [@desmedt2022measures; @vlemincx2023novel]. Ultimately, these studies show that the IAS is most informative when treated as a set of domain-specific, context-sensitive indicators rather than as a broad global index, advancing both the measurement and conceptual understanding of interoception.
